### 2020-07-28 manuscript.tex
fig1 update

The molecular clock hypothesis states that the evolutionary rates of biological sequences are approximately constant throughout time \cite{zuckerkandl1962molecular}.
This assumption forms the basis of phylogenetics, under which evolutionary trees and divergence dates of life forms are inferred from biological sequences, such as nucleotides and amino acids \cite{douzery2003local, drummond2006relaxed}.
In Bayesian phylogenetics, these trees and their associated parameters are estimated as probabilitiy distributions \cite{kuhner1995estimating, larget1999markov, mau1999bayesian}. 
Statistical inference is typically performed by sampling under the Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm \cite{metropolis53, hastings70}.
Bayesian phylogenetic clock models have been implemented in many platforms including BEAST \cite{drummond2012bayesian}, BEAST2 \cite{bouckaert2019beast}, MrBayes \cite{ronquist2012mrbayes}, and RevBayes \cite{hohna2016revbayes}.

@@

With the increasing availabilty of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational runtime but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels including the Bactrian and mirror kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of clock models, smart tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

@@

\begin{eqnarray}
\label{eq:bayesian}
p(\mathcal{T}, \vec{\mathcal{R}}^{\,}, \sigma, \theta|D) \propto  p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \theta) \; p(\mathcal{T}|\theta) \;  p(\vec{\mathcal{R}}^{\,} | \sigma) \; p(\sigma) \; p(\theta),
\end{eqnarray}

where $\sigma$ is the clock model standard deviation, $p(\mathcal{T}|\theta)$ is the tree prior, and $\theta$ describes other unspecified parameters. 
The tree likelihood $p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \theta)$ is computed using Felsenstein's tree-peeling algorithm \cite{felsenstein1981evolutionary}, where
$\vec{\mathcal{R}}^{\,}$ is a vector of abstracted substitution rates and transformed into real rates by function $r(\vec{\mathcal{R}}^{\,})$. Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

Let $t_i$ be the height (time) of node $i$.
Every internal and leaf node $i$ in $\mathcal{T}$ is associated with a parental branch length (the height difference between $i$ and its parent) $\tau_i$ and a parental branch substitution rate $r_i = r(\mathcal{R}_i)$. 
There are a total of $2N-2$ internal and leaf nodes. 
Under the relaxed clock model, each $\mathcal{R}_i$ is independently distributed under the prior.

The posterior distribution is sampled by MCMC.
The probability of accepting proposed state $x^\prime$ from state $x$ is calculated using the Metropolis-Hastings ratio \cite{metropolis53, hastings70}:

\begin{eqnarray}
\label{eq:MCMC}
\alpha(x^\prime|x) =  \min\left( 1, \frac{p(x^\prime|D)}{p(x|D)} \frac{q(x|x^\prime)}{q(x^\prime|x)} \right).
\end{eqnarray}

$q(a|b)$ is the transition kernel: the probability of proposing state $b$ from state $a$.
The ratio between the two $\frac{q(x|x^\prime)}{q(x^\prime|x)}$ is known as the Hastings ratio \cite{hastings70}.

@@

In Bayesian inference, the way parameters are represented in the model can affect the mixing ability of the model and the meaning of the model itself \cite{gelman2004parameterization}. Three methods for parameterising substitution rates are described below, and are later evaluated in \textbf{\nameref{sect:results}}. Each parameterisation technique is associated with i) an abstraction of the rates $\vec{\mathcal{R}}^{\,}$, ii) some function for transforming this parameter into real rates $r(\vec{\mathcal{R}}^{\,})$, and iii) a prior density function of the abstraction $p(\vec{\mathcal{R}}^{\,} | \sigma) $. The three methods are summarised in \textbf{Fig \ref{fig:rateparams}}.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Methods of rate parameterisation.} The \textit{cat} and \textit{quant} approximations are plotted on top of the true underlying rate prior distribution (\textit{real}). In this example, rates are drawn from a $\text{LogNormal}(\mu = -0.045, \sigma = 0.3)$ distribution. The probability density function (pdf) and cumulative density function (cdf) of this distribution are shown.}
\label{fig:rateparams}
\end{figure}

@@

We hypothesis that datasets with strong signal (or large $L$) will mix best when the more precise and meticulous kind of operator is employed, such as those informed by correlations within the posterior distribution e.g. the family of constant distance operators \cite{zhang2020improving}.
Whereas, datasets which contain very poor signal (or small $L$) are likely to mix best when there is more weight placed on bold operators such as \texttt{SampleFromPrior}, \texttt{Swap}, or \texttt{Uniform} which sample each branch rate independently of its current estimate.
\texttt{AdaptiveOperatorSampler} should learn the combination of weights behind these classes of operators best suited for the dataset.

@@

Methodologies are benchmarked using one simulated and nine empirical datasets -- the latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear et al. as `benchmark alignments' (\textbf{Table \ref{table:datasets}}).
To facilitate convergence on complex datasets, each posterior distribution is sampled using coupled MCMC (MC$^3$) with 4 chains \cite{muller2019coupled}.
Each methodology is benchmarked across 5 replicates on each dataset, using an the Intel Xeon Gold 6138 CPU (2.00 GHz).

@@

% https://github.com/roblanf/BenchmarkAlignments
\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l l l|} 
 \hline
  & $N$ & $P$ & $L$ (kb) & $L_\text{eff}$ (kb) & \textbf{Description} \\

@@

\clearpage
\subsection*{Round 1: A simple operator weight learning algorithm can improve performance}

@@

\end{tabular}
\caption{Operator configurations for Round 1. In each configuration, the weight behind $\mathcal{R}$ sums to 30 and for $\sigma$ is equal to 10 within any one rate parameterisation. Operators which apply to either internal nodes or the root (but not both) weighted according to leaf count $N$.}
\label{table:round1Results}
\end{table}

>>>>

The molecular clock hypothesis states that the evolutionary rates of biological sequences are approximately constant throughout time \cite{zuckerkandl1962molecular}.
This assumption forms the basis of phylogenetics, under which evolutionary trees and divergence dates of life forms are inferred from biological sequences, such as nucleotides and amino acids \cite{douzery2003local, drummond2006relaxed}.
In Bayesian phylogenetics, these trees and their associated parameters are estimated as probabilitiy distributions \cite{kuhner1995estimating, larget1999markov, mau1999bayesian}. 
Statistical inference is typically performed by sampling using the Markov chain Monte Carlo (MCMC) algorithm \cite{metropolis53, hastings70}.
Bayesian phylogenetic clock models have been implemented in many platforms including BEAST \cite{drummond2012bayesian}, BEAST2 \cite{bouckaert2019beast}, MrBayes \cite{ronquist2012mrbayes}, and RevBayes \cite{hohna2016revbayes}.

@@

With the increasing availabilty of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational runtime but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels such as the Bactrian and mirror kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of clock models, informed tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

@@

\begin{eqnarray}
\label{eq:bayesian}
p(\mathcal{T}, \vec{\mathcal{R}}^{\,}, \sigma, \theta|D) \propto  p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \theta) \; p(\mathcal{T}|\theta) \;  p(\vec{\mathcal{R}}^{\,} | \sigma) \; p(\sigma) \; p(\theta).
\end{eqnarray}

$\sigma$ represents clock model related parameters, and $p(\mathcal{T}|\theta)$ is the tree prior where $\theta$ describes further unspecified parameters. 
The tree likelihood $p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \theta)$ is computed using the tree-peeling algorithm \cite{felsenstein1981evolutionary}, where
$\vec{\mathcal{R}}^{\,}$ is a vector of abstracted substitution rates and transformed into real rates by function $r(\vec{\mathcal{R}}^{\,})$. Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

Let $t_i$ be the height (time) of node $i$.
Each node $i$ in $\mathcal{T}$, except for the root, is associated with a parental branch length $\tau_i$ (the height difference between $i$ and its parent)  and a parental branch substitution rate $r_i = r(\mathcal{R}_i)$. 
In a relaxed clock model, each of the $2N-2$ elements in $\mathcal{R}_i$ are independently distributed under the prior $p(\vec{\mathcal{R}}^{\,} | \sigma)$.

The posterior distribution is sampled by the Metropolis-Hastings-Green MCMC algorithm \cite{metropolis53, hastings70, green1995reversible},
under which the probability of accepting proposed state $x^\prime$ from state $x$ is equal to:

\begin{eqnarray}
\label{eq:MCMC}
\alpha(x^\prime|x) =  \min\left( 1, \frac{p(x^\prime|D)}{p(x|D)} \; \frac{q(x|x^\prime)}{q(x^\prime|x)} \; |J| \right).
\end{eqnarray}

$q(a|b)$ is the transition kernel: the probability of proposing state $b$ from state $a$.
The ratio between the two $\frac{q(x|x^\prime)}{q(x^\prime|x)}$ is also known as the Hastings ratio \cite{hastings70}.
The determinant of the Jacobian matrix $|J|$ solves the dimension-matching problem for proposals which operate on multiple terms across one or more spaces \cite{green1995reversible, geyer2003metropolis}. 
This term is known as the Green ratio.

@@

In Bayesian inference, the way parameters are represented in the model can affect the mixing ability of the model and the meaning of the model itself \cite{gelman2004parameterization}. Three methods for parameterising substitution rates are described below.
 Each parameterisation is associated with i) an abstraction of the branch rate vector $\vec{\mathcal{R}}^{\,}$, ii) some function for transforming this parameter into unabstracted branch rates $r(\vec{\mathcal{R}}^{\,})$, and iii) a prior density function of the abstraction $p(\vec{\mathcal{R}}^{\,} | \sigma) $. 
 The three methods are summarised in \textbf{Fig \ref{fig:rateparams}}.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the branch rate prior under a log-normal($-0.5\sigma^2, \sigma$) distribution (with a mean of 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For simplicity there are 10 \textit{cat} bins displayed, however in practice there are $2N-2$ bins. }
\label{fig:rateparams}
\end{figure}

@@

We hypothesise that datasets with strong signal (or large $L$) will mix best when the more precise and meticulous kind of operator is employed, such as those informed by correlations within the posterior distribution e.g. the family of constant distance operators \cite{zhang2020improving}.
Whereas, datasets which contain very poor signal (or small $L$) are likely to mix best when there is more weight placed on bold operators such as \texttt{SampleFromPrior}, \texttt{Swap}, or \texttt{Uniform} which sample each branch rate independently of its current estimate.
\texttt{AdaptiveOperatorSampler} should learn the combination of weights behind these classes of operators best suited for the dataset.

@@

Methodologies are benchmarked using one simulated and nine empirical datasets -- the latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear et al. as `benchmark alignments' (\textbf{Table \ref{table:datasets}}).
To facilitate convergence on complex datasets, each posterior distribution is sampled using coupled MCMC (MC$^3$) with 4 chains \cite{muller2019coupled}.
Each methodology is benchmarked across 5 replicates on each dataset, using an the Intel Xeon Gold 6138 CPU (2.00 GHz) and all methodologies use identical models and operator configurations, except where a difference is specified.

@@

% https://github.com/roblanf/BenchmarkAlignments
\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l l l|} 
 \hline
  & $N$ & $P$ & $L$ (kb) & $L_\text{eff}$ (kb) & \textbf{Description} \\
  \hline

@@

\clearpage
\subsection*{Round 1: A simple operator-weight learning algorithm can improve performance}

@@

\end{tabular}
\caption{Operator configurations for Round 1. In each configuration, the weight behind $\mathcal{R}$ sums to 30 and for $\sigma$ is equal to 10 within any one rate parameterisation. Operators which apply to either internal nodes or the root (but not both) weighted according to leaf count $N$. The adapt operators are further broken down in \textbf{Table \ref{table:adaptiveSampling}}.}
\label{table:round1Results}
\end{table}
